---
title: package comparison
date: "`r Sys.time()`"
---

<!-- http://zevross.com/blog/2014/07/09/making-use-of-external-r-code-in-knitr-and-r-markdown/ -->

```{r pkgs,message=FALSE}
## fitting methods
library(lme4)
library(blme)
library(glmmTMB)
library(brms)
## plotting/manipulation
library(broom)  ## install_github("bbolker/broom")
library(dotwhisker)
library(dplyr)
library(ggplot2)
```

```{r read_chunk,echo=FALSE}
knitr::read_chunk("bglmer_runs.R")
```

Define a formula (which we'll use repeatedly) and make a data frame that represents a fully crossed, randomized-block design with three factors for the fixed effects (3x3x2) and two random effects (`id` and `item.new`).

```{r setup_runs,eval=FALSE}
```

Simulate a Bernoulli response:

```{r simulate_data,eval=FALSE}
```

Fit the model with `lme4`, `blme`, `glmmTMB`, and `brms` (the first and third are un-penalized, the second and fourth set a zero-centered prior on the fixed effects):

```{r run_models,eval=FALSE}
```

Retrieve the stored data:
```{r get_data}
resList <- readRDS("bglmer_runs.rds")
```

Tweak labels a bit and put the results together:
```{r fix_data,warning=FALSE,echo=FALSE}
## lme4 non-pos-def Hessian warnings suppressed ...
tt <- lapply(resList,tidy)
## fix brms term names
tt$brms <- tt$brms %>% filter(grepl("^(b|sd)_",term)) %>%
    mutate(term=gsub("^b_","",term),
           term=gsub("__",".",term),
           term=gsub("([a-z\\.]+)\\.Intercept","Intercept.\\1",term),
           term=gsub("Intercept","(Intercept)",term))
resFrame <- bind_rows(tt,.id="model")
```

Plot:

```{r do_plot}
dwplot(resFrame)+theme_bw()+
    geom_vline(xintercept=1,lty=2)+
    scale_x_continuous(limits=c(-5,5),oob=scales::squish)+
    scale_colour_brewer(palette="Set1")
```

Conclusions:

- all approaches give generally similar answers
- even this fairly large data set isn't enough to give accurate answers about interactions (unpenalized models give ridiculous answers)
- it's a bit worrying that 2/12 parameters aren't covered here (i.e., confidence intervals for `c.con.tr3` and `c.con.tr3:c.diff.tr2`); this could be a weird single simulation (I don't have the patience to redo this hundreds of times - `brms` in particular runs slowly)

### Session info

```{r si}
sessionInfo()
```
