---
author: Ben Bolker
title: "efficiently fitting many mixed models with different fixed effects in `lme4`"
---
```{r pkg,message=FALSE}
library("lme4")
```

```{r sim}
set.seed(101)
genotypes <- c("A","B","C")
n <- 1000
ngenes <- 100
## sample cluster sizes up to a cumulative total of 1000
clustsize <- 1+rpois(500,lambda=1)
clust <- rep(1:500,clustsize)[1:n]
dd <- data.frame(fgluc=rnorm(n),
                 clust=factor(clust))
genes <- setNames(do.call(data.frame,
                 replicate(ngenes,sample(
                     x=genotypes,size=n,replace=TRUE),
                           simplify=FALSE)),
                  paste0("gene",1:ngenes))
dd2 <- cbind(dd,genes)
```


```{r}
fitGene <- function(g) {
    f <- reformulate(c(g,"(1|clust)"),response="fgluc")
    lmer(f,data=dd2)
}
```

See `?modular` for the basic idea.
here we are taking the pre-computed information about the model
structure (in particular the 'reTrms' component which describes
the RE structure and involves an expensive permutation computation)
and filling in the new formula and fixed-effect model matrix for
each new predictor variable.
I thought we might save some more time by returning just the
optimized parameters and not the whole model structure, but that
apparently that doesn't make much difference ...
It might be possible to copy the existing structures at an even
lower level (and hence gain a little bit more speed), but this
is very low-hanging fruit ...

```{r refitgene}
lmod0 <- lFormula(fgluc ~ 1 + (1|clust), dd2)
refitGene <- function(g,retmod=TRUE) {
    f0 <- reformulate(g,response="fgluc")
    f <- reformulate(c(g,"(1|clust)"),response="fgluc")
    lmod <- lmod0
    lmod$formula <- f
    lmod$X <- model.matrix(f0,data=dd2)
    devfun <- do.call(mkLmerDevfun, lmod)
    opt <- optimizeLmer(devfun)
    if (!retmod) opt else {
        mkMerMod(environment(devfun), opt, lmod$reTrms, fr = lmod$fr)
    }
}
```

```{r}
fitAll <- function(...) {
    lapply(grep("^gene",names(dd2),value=TRUE),...)
}
```

```{r}
library(rbenchmark)
benchmark(fitAll(fitGene),
          fitAll(refitGene),
          fitAll(refitGene,retmod=FALSE),
          replications=10)
```
