---
title: "estimating group/multi-membership models with lme4"
author: "Ben Bolker"
date: "`r format(Sys.time(),'%d %B %Y')`"
---

Consider a situation where we have multiple items of some sort which have associated effects, but we can't association observations with single items; instead, each observation is associated with a *group* of items. (Two examples which have come up are (1) authorship of papers and (2) hockey players on a team.) These are *not* quite identical to "multi-membership models" (I think), although similar techniques to those shown here could work for multi-membership models.

Load packages (we don't really need anything beyond `lme4`; the rest are for convenience/drawing pictures).

```{r pkgs,message=FALSE}
library(lme4)
library(broom.mixed)
library(ggplot2); theme_set(theme_bw())
library(Matrix)
```

Construct a simulated example: first, simulate the
design (structure).

```{r simdesign}
nm <- 20
nobs0 <- 500
set.seed(101)
## choose items for observations
W <- matrix(rbinom(nobs0*nm, prob=0.25, size=1),nrow=nobs0,ncol=nm)
dimnames(W) <- list(NULL,LETTERS[seq(nm)])
W[1:5,]
table(rowSums(W))
```

The first 10 observations:

```{r showobs}
image(Matrix(W),ylim=c(1,10),sub="",ylab="Observation",
      xlab="Item",
      ## draw tick labels at top
      scales=list(at=1:20,x=list(labels=colnames(W),
                                 alternating=2)))
```

Since we chose which items/individuals to include for each observation
randomly and independently (Bernoulli with probability 0.25),
we have a highly variable number of individuals present
for different observations (1-11).
This would be realistic for some examples (authorship), unrealistic
for others (hockey) ... I don't think it really makes much difference
computationally (statistically, having some observations with a single
member must make estimation more powerful ...)
The 0/1 matrix (indicator variable for whether item $i$ is included
in observation $j$) is convenient, and will turn out to be the form
we need for inclusion in the model. It should be fairly straightforward
to convert other forms (e.g. a list of sets of items associated with
each observation) to this form ...

Now simulate the response variable.

```{r simvals}
b <- rnorm(nm)  ## item-level effects
beta <- c(1,2)
M <- data.frame(x=runif(nobs0))
X <- model.matrix(~x, data=M)
## n.b. get in trouble if we don't add residual error
## (theta is scaled relative to residual error)
## here, use theta=sigma=1
M$y <- c(X %*% beta) + c(W %*% b) +rnorm(nobs0,sd=1)
```

Fit (see `?modular`):
```{r modular_fit}
## helpful to specify a factor with the right levels:
## actual values are unimportant since we will specify Zt/Ztlist directly
M$fake <- rep(LETTERS[seq(nm)],length.out=nobs)
lmod <- lFormula(y~x+(1|fake), data=M)
lmod$reTrms$Zt <- lmod$reTrms$Ztlist[[1]] <- Matrix(t(W))
devfun <- do.call(mkLmerDevfun, lmod)
opt <- optimizeLmer(devfun)
m1 <- mkMerMod(environment(devfun), opt, lmod$reTrms, fr = lmod$fr)
```

Results look OK (correct fixed effects, item and residual variance estimated):

```{r results}
m1
```

The conditional modes by item look good:

```{r plotresults}
dd <- tidy(m1, effects="ran_vals")
dd <- transform(dd, level=reorder(level,estimate))
truth <- data.frame(level=LETTERS[seq(nm)],estimate=b)
ggplot(dd,aes(x=level,y=estimate))+
    geom_pointrange(aes(ymin=estimate-2*std.error,
              ymax=estimate+2*std.error))+coord_flip()+
    geom_point(data=truth,colour="red")
```

## see also

```{r findFn,eval=FALSE}
sos::findFn("{multiple membership}") ## GPvam, R2MLwiN, RealVAMS, ...
```

- `r-sig-mixed-models` posts ([here](https://stat.ethz.ch/pipermail/r-sig-mixed-models/2012q1/017826.html), [here](https://stat.ethz.ch/pipermail/r-sig-mixed-models/2011q2/006318.html))
- `mm()`, `mmc()` in the `brms` package
- `mult.memb()` in the `MCMCglmm()` package

## function encapsulation

Suppose we have a real data frame `M` and a (sparse) weight matrix `W` (or? a list `L` of weight vectors?

```{r}
##' @param formula mixed model formula
##' @param data data frame (possibly but not necessarily containing factors
##' @param memb_mat list of weights matrices  with which to replace Zt components
lmer_multimemb <- function(formula,data,memb_mat=list(),...) {
  ## FIXME: pass ... through appropriately
  ## FIXME: test dimensions
  mnms <- names(memb_mat)
  fb <- findbars(formula)
  gvars <- vapply(fb, function(x) deparse(x[[3]]), character(1))
  Ztlist <- list()
  for (i in seq_along(fb)) {
    fbnm <- deparse(fb[[i]])
    ## find corresponding random-effects term
    w <- which(mnms==gvars[i])
    if (length(w)>0) {
      M <- Matrix::Matrix(memb_mat[[w]])
      ## extract LHS (effect)
      form <- as.formula(substitute(~z,list(z=fb[[i]][[2]])))
      ## construct model matrix & compute Khatri-Rao product 
      X <- model.matrix(form,data=data)
      Zt <- Matrix::KhatriRao(t(M),t(X),make.dimnames=TRUE)
      ## FIXME: mess with names?
      Ztlist[[fbnm]] <- Zt
      ## if necessary, add factor to data
      if (!m %in% names(data)) {
        ## if the factor has non-trivial ordering, it should be included
        ## in the data.  Do we have to worry about ordering of Z? test!
        data[[gvars[i]]] <- factor(colnames(memb_mat[[w]]))
      }
    } ## if  (length(w)>0)
  } ## for i in seq(fb)
  ## call lFormula  (FIXME: allow glFormula)
  lmod <- lFormula(formula,data=data)
  ## substitute new Ztlist elements
  for (m in names(Ztlist)) {
    lmod$reTrms$Ztlist[[m]] <- Ztlist[[m]]
  }
  lmod$reTrms$Zt <- do.call(rbind,lmod$reTrms$Ztlist)
  ## finish fitting
  devfun <- do.call(mkLmerDevfun, lmod)
  opt <- optimizeLmer(devfun)
  m1 <- mkMerMod(environment(devfun), opt, lmod$reTrms, fr = lmod$fr)
  return(m1)
}
```

This seems to work ...

```{r test}
lmer_multimemb(formula=y ~ x + (1|team),
               data=M,
               memb_mat=list(team=W))
```

## Data formats

We want to allow users to conveniently transform whatever data formats they have to a sparse incidence matrix.

### memberships in multiple columns

Maybe it's in multiple columns (as in this example from Harold Doran):

```{r doran_dat}
tmp <- data.frame(id = 1:5,
                  tch1 = c(10,10,11,11,12),
                  tch2 = c(77,15,NA,10,11),
                  tch3 = c(NA, NA, 77, 12, 10),
                  y = rnorm(5), x = rnorm(5))
## list of teachers
all_tch <- with(tmp,sort(unique(na.omit(c(tch1,tch2,tch3)))))
## empty (sparse) matrix with dimnames
W <- Matrix(0,nrow=nrow(tmp),ncol=length(all_tch),
            dimnames=list(as.character(tmp$id),as.character(all_tch)))
## melt all teacher columns to long/combined format
all_prs0 <- na.omit(reshape2::melt(tmp[,1:4],id="id"))
all_prs1 <- all_prs0[,c("id","value")]
## convert to numeric indices
all_prs2 <- all_prs1
all_prs2[] <- lapply(all_prs2,function(x) as.numeric(factor(x)))
## convert to two-column matrix
all_prs2 <- as.matrix(all_prs2)
W[all_prs2] <- 1 ## weights
```

## row_ID, col_ID, weight as three columns

As pointed out by Robyn Wimmers, `reshape2::dcast` is handy if your data are in a three-column format (row_label, col_label, weight). It's marginally less efficient because it initially generates a dense matrix.

```{r threecol}
## convert from the format we have above
all_prs3 <- all_prs1
all_prs3[] <- lapply(all_prs3, factor)
all_prs3$wts <- 1
## now data frame -> matrix
W0 <- reshape2::dcast(all_prs3,id~value, value.var="wts", fill=0)
idvars <- W0$id
W <- Matrix(as.matrix(W0[,-1])) ## drop 'id' column and sparsify
rownames(W) <- as.character(idvars)
```

---

Session info:
```{r SI}
sessionInfo()
```

