---
title: "estimating group membership models with lme4"
author: "Ben Bolker"
date: "`r format(Sys.time(),'%d %B %Y')`"
---

Consider a situation where we have multiple items of some sort which have associated effects, but we can't association observations with single items; instead, each observation is associated with a *group* of items. (Two examples which have come up are (1) authorship of papers and (2) hockey players on a team.) These are *not* quite identical to "multi-membership models" (I think), although similar techniques to those shown here could work for multi-membership models.

Load packages (we don't really need anything beyond `lme4`; the rest are for convenience/drawing pictures).

```{r pkgs,message=FALSE}
library(lme4)
library(broom.mixed)
library(ggplot2); theme_set(theme_bw())
library(Matrix)
```

Construct a simulated example: first, simulate the
design (structure).

```{r simdesign}
nm <- 20
nobs <- 500
set.seed(101)
## choose items for observations
pres <- matrix(rbinom(nobs*nm, prob=0.25, size=1),nrow=nobs,ncol=nm)
dimnames(pres) <- list(NULL,LETTERS[seq(nm)])
pres[1:5,]
table(rowSums(pres))
```

The first 10 observations:

```{r showobs}
image(Matrix(pres),ylim=c(1,10),sub="",ylab="Observation",
      xlab="Item",
      ## draw tick labels at top
      scales=list(at=1:20,x=list(labels=colnames(pres),
                                 alternating=2)))
```

Since we chose which items/individuals to include for each observation
randomly and independently (Bernoulli with probability 0.25),
we have a highly variable number of individuals present
for different observations (1-11).
This would be realistic for some examples (authorship), unrealistic
for others (hockey) ... I don't think it really makes much difference
computationally (statistically, having some observations with a single
member must make estimation more powerful ...)
The 0/1 matrix (indicator variable for whether item $i$ is included
in observation $j$) is convenient, and will turn out to be the form
we need for inclusion in the model. It should be fairly straightforward
to convert other forms (e.g. a list of sets of items associated with
each observation) to this form ...

Now simulate the response variable.

```{r simvals}
b <- rnorm(nm)  ## item-level effects
## n.b. get in trouble if we don't add residual error
## (theta is scaled relative to residual error)
## here, use theta=sigma=1
y <- c(pres %*% b) +rnorm(nobs,sd=1)
```

Fit (see `?modular`):
```{r modular_fit}
## helpful to specify a factor with the right levels:
## actual values are unimportant since we will specify Zt/Ztlist directly
fake <- rep(LETTERS[seq(nm)],length.out=nobs)
lmod <- lFormula(y~1+(1|fake))
lmod$reTrms$Zt <- lmod$reTrms$Ztlist[[1]] <- Matrix(t(pres))
devfun <- do.call(mkLmerDevfun, lmod)
opt <- optimizeLmer(devfun)
m1 <- mkMerMod(environment(devfun), opt, lmod$reTrms, fr = lmod$fr)
```

Results look OK (correct item and residual variance estimated):

```{r results}
m1
```

The conditional modes by item look good:

```{r plotresults}
dd <- tidy(m1, effects="ran_vals")
dd <- transform(dd, level=reorder(level,estimate))
truth <- data.frame(level=LETTERS[seq(nm)],estimate=b)
ggplot(dd,aes(x=level,y=estimate))+
    geom_pointrange(aes(ymin=estimate-2*std.error,
              ymax=estimate+2*std.error))+coord_flip()+
    geom_point(data=truth,colour="red")
```

See also: 

---

Session info:
```{r SI}
sessionInfo()
```

