---
title: "Phylogenetic comparative methods in the lme4-verse"
author: Ben Bolker
date: "`r format(Sys.time(), '%H:%M %d %B %Y')`"
output:
    html_document:
    mathjax: default
---

```{r pkgs,message=FALSE}
library(ape)
library(lme4)
library(Matrix)
## additional possibilities for PGLMMs ...
library(MCMCglmm)
library(MASS) ## glmmPQL + ape::corBrownian()
library(pez)
## n.b. at present need 'modular' devel branch:
## devtools::install_github("glmmTMB/glmmTMB/glmmTMB",ref="modular")
library(glmmTMB)
## utils
library(dplyr)
library(coda)
library(lattice)
library(broom) ## need install_github("bbolker/broom")
library(dotwhisker)
```

- The standard problem of *phylogenetic comparative methods* is to analyze relationships among data where the observations are gathered from nodes (usually tips) of a phylogenetic tree - for example, regression analyses of body temperature as a function of body size for animals within a clade
- More generally, we can frame this in the usual GLMM way as
$$
\begin{split}
y & \sim D(\mu,\phi) \\
\mu & = g^{-1}(\eta) = g^{-1}(X \beta + Z b) \\
b & \sim \textrm{MVN}(0,\Sigma)
\end{split}
$$
where the part that makes it specifically phylogenetic is that $\Sigma$ captures the *phylogenetic correlation*. The PC is the correlation among observations due to relatedness; recently diverged taxa have higher correlation than more anciently diverged taxa. In the extreme case of a *star phylogeny* (all taxa diverged from each other simultaneously at some point in the past) the phylogenetic correlation collapses to a diagonal matrix and we get back to the simple, uncorrelated regression.

Various P(G)LMM (phylogenetic [generalized] linear mixed model] approaches have been proposed. Many depend on Pagel's lambda transformation, which gives the correlation matrix a particularly simple form (but has been [criticized](http://www.carlboettiger.info/2013/10/11/is-it-time-to-retire-pagels-lambda.html) ...)

An alternative approach is to model the phylogenetic correlation as a *Gaussian process*. In particular, suppose that the evolutionary process is a Brownian motion (an almost certainly incorrect/oversimplified model of evolution, but one that many phylogenetic methods are built on). In that case, the phylogenetic variability of a particular observation can be written as the sum of the evolutionary changes that occurred on all of the branches in the phylogeny in its past. If we set up the $Z$ matrix appropriately, we can model everything with a sequence of *independent* errors, rather than having to do fancy things to impose a correlation structure on the random effects.

## Nuts and bolts: from a phylogeny to a $Z$ matrix for the GP

```{r setup}
library(ape)
set.seed(101)
r <- makeNodeLabel(rtree(4))
plot(r,show.node.label=TRUE)
```

Information in a `phylo` object is contained in the *edge matrix*:

>    edge: a two-column matrix of mode numeric where each row represents
          an edge of the tree; the nodes and the tips are symbolized
          with numbers; the tips are numbered 1, 2, ..., and the nodes
          are numbered after the tips. For each row, the first column
          gives the ancestor.

```{r edge}
t(r$edge)
```

and a list of *edge lengths*

```{r edgelen}
r$edge.length
```

Inspecting this tree, we can figure out (see `$tip.label` and `$node.label` for label-to-number correspondences):

- tips are 1-4, nodes are 5-7
- tip 1 (`t1`) involves branches 2 (6 $\to$ 1) and 1 (5 $\to$ 6).
- tip 2 (`t3`) involves branches 3 (6 $\to$ 2) and 1 (5 $\to$ 6)
- tip 3 (`t2`) involves branches 5 (7 $\to$ 3) and 4 (5 $\to$ 7)
- tip 4 (`t4`) involves branches 6 (7 $\to$ 4) and 4 (5 $\to$ 7)

So, for example, we can say that the 'error' value corresponding to tip 1 is
$\ell_1 b_1 + \ell_2 b_2$,  where $\ell_i$ is the (square root of??) the branch length
and the $b_i$ are independent, homoscedastic Normal variates. Alternately, the $Z$ matrix is

$$
\begin{pmatrix}
\ell_1 & \ell_2 &   0     & 0 & 0 & 0 \\
\ell_1 &    0   &  \ell_3 & 0 & 0 & 0 \\
0 & 0 &  0  & \ell_4 & \ell_5 & 0 \\
0 & 0 &  0  & \ell_4 & 0 & \ell_6
\end{pmatrix}
$$
where $\ell_i$ is the length of the $i^\textrm{th}$ branch,
so that the species effects are $Z b$.

If we can build the corresponding $Z$ matrix, then we can insert it in
the `lme4` modular model-fitting process (see `?modular`).

Here's a (probably not very efficent) way to construct the Z matrix.
(There must be a way to not walk the tree multiple times from every tip ...

```{r phylo.to.Z}
phylo.to.Z <- function(r) {
    ntip <- length(r$tip.label)
    Z <- Matrix(0.0,ncol=length(r$edge.length),nrow=ntip)
    nodes <- (ntip+1):max(r$edge)
    root <- nodes[!(nodes %in% r$edge[,2])]
    for (i in 1:ntip) {
        cn <- i  ## current node
        while (cn != root) {
            ce <- which(r$edge[,2]==cn)   ## find current edge
            Z[i,ce] <- r$edge.length[ce]  ## set Z to branch length
            cn <- r$edge[ce,1]            ## find previous node
        }
    }
    return(Z)
}
phylo.to.Z(r)
```

```{r phylo.to.Z2}
```

(This could benefit from the repeated-entry sparse matrix class that Steve Walker wrote.)

On the other hand, it only takes a few seconds to run for a 200-species phylogeny (see below).

## constructing a GP PGLMM with lme4: machinery

"All" we need to do is (1) call `(g)lFormula`, with a formula that includes a `(1|phylo)` term, to build the basic (wrong) structure; (2) modify the `reTrms` component of the structure appropriately; (3) go through the rest of the modular procedure for building a (G)LMM.

```{r phylocode}
#' split a square (block) matrix into component blocks 
#' @param M square matrix
#' @param ind indices (0,n1,n2,...) giving the endpoint of each block
split_blkMat <- function(M,ind) {
    res <- list()
    if (length(ind)==1) return(list(M))
    for (i in 1:(length(ind)-1)) {
        v <- (ind[i]+1):ind[i+1]
        res[[i]] <- M[v,v]
    }
    return(res)
}

#' modify reTrms object
#' @param rt a reTrms object
#' @param phylo a phylo object (phylogenetic tree)
#' @param phylonm name of phylogenetic term in model
#' @param phyloZ Z matrix built on branch length
modify_phylo_retrms <- function(rt,phylo,phylonm="phylo",
                                phyloZ=phylo.to.Z(phylo)) {
    ## FIXME: better way to specify phylonm
    ## need to replace Zt, Lind, Gp, flist, Ztlist
    ## we have the same number of parameters (theta, lower),
    ##  same number of obs
    n.edge <- nrow(phylo$edge)
    phylo.pos <- which(names(rt$cnms)==phylonm)
    inds <- c(0,cumsum(sapply(rt$Ztlist,nrow)))
    ## Zt: substitute phylo Z for previous dummy (scalar-intercept) Z
    rt[["Ztlist"]][[phylo.pos]] <- t(phyloZ)
    ## reconstitute Zt from new Ztlist
    rt[["Zt"]] <- do.call(rbind,rt[["Ztlist"]])
    ## Gp: substitute new # random effects (n.edge) for old # (n.phylo)
    Gpdiff <- diff(rt$Gp)  ## old numbers
    Gpdiff_new <- Gpdiff
    Gpdiff_new[phylo.pos] <- n.edge  ## replace
    rt[["Gp"]] <- as.integer(c(0,cumsum(Gpdiff_new)))          ## reconstitute
    ## Lind: replace phylo block with the same element, just more values
    Lind_list <- split(rt[["Lind"]],rep(seq_along(Gpdiff),Gpdiff))
    Lind_list[[phylo.pos]] <- rep(Lind_list[[phylo.pos]][1],n.edge)
    rt[["Lind"]] <- unlist(Lind_list)
    ## Lambdat: replace block-diagonal element in Lambdat with a
    ##   larger diagonal matrix
    Lambdat_list <- split_blkMat(rt[["Lambdat"]],inds)
    Lambdat_list[[phylo.pos]] <- Diagonal(n.edge,1.0)
    rt[["Lambdat"]] <- Matrix::.bdiag(Lambdat_list)
    ## flist: 
    rt[["flist"]] <- as.list(rt[["flist"]])
    rt[["flist"]][[phylonm]] <- factor(paste0("edge_",seq(n.edge)))
    return(rt)
}

#' 
phylo_glmm <- function(formula,data,family,phylo,phyloZ) {
    glmod <- glFormula(formula=formula,data = data, family = family)
    glmod$reTrms <- modify_phylo_retrms(glmod$reTrms,phylo,
                                        phylonm="phylo",phyloZ)
    devfun <- do.call(mkGlmerDevfun, glmod)
    opt <- optimizeGlmer(devfun)
    devfun <- updateGlmerDevfun(devfun, glmod$reTrms)
    opt <- optimizeGlmer(devfun, stage=2)
    mkMerMod(environment(devfun), opt, glmod$reTrms, fr = glmod$fr)
}

phylo_lmm <- function(formula,data,phylo,phyloZ,control) {
  lmod <- lFormula(formula=formula,data = data,control=control)
  lmod$reTrms <- modify_phylo_retrms(lmod$reTrms,phylo,
                                     phylonm="phylo",phyloZ)
  devfun <- do.call(mkLmerDevfun, lmod)
  opt <- optimizeLmer(devfun)
  # devfun <- updateLmerDevfun(devfun, lmod$reTrms)
  # opt <- optimizeLmer(devfun, stage=2)
  mkMerMod(environment(devfun), opt, lmod$reTrms, fr = lmod$fr)
}
```

## glmmTMB fits: nuts and bolts

`glmmTMB` can be deconstructed in a similar way. In fact, we can re-use a lot of the
machinery. Being able to use `glmmTMB` means we can use a broader range of distributions,
zero-inflation, etc. (machinery below assumes phylogenetic structure only in the
conditional distribution). This is also a little clunky, some adjustment on the `glmmTMB`
side might make it a bit easier.

```{r tmbfit0}
modify_TMBstruc <- function(tmbstruc,phylo,phylonm="phylo",
                            phyloZ=phylo.to.Z(phylo)) {
    n.edge <- nrow(phylo$edge)
    ## stuff in tmbstruc (but not within data.tmb) is (maybe) necessary
    ##  for cosmetics, proper reporting
    tmbstruc$condList$reTrms <- modify_phylo_retrms(tmbstruc$condList$reTrms,
                                                    phylo,phylonm,phyloZ)
    tmbstruc$condReStruc$`1 | phylo`$blockReps <- n.edge
    tmbstruc$condList$Z <- t(tmbstruc$condList$reTrms$Zt)
    ## data *inside* data.tmb is actually the most critical to allow correct fit
    tmbstruc$data.tmb$terms$`1 | phylo`$blockReps <- n.edge
    tmbstruc$data.tmb$Z <- t(tmbstruc$condList$reTrms$Zt)
    tmbstruc$parameters$b <- rep(0,ncol(tmbstruc$data.tmb$Z))
    return(tmbstruc)
}
```


## example

### get data

From [chapter 11](http://www.mpcm-evolution.org/practice/online-practical-material-chapter-11/chapter-11-1-simple-model-mcmcglmm) of Garamszegi (ed.): data are [here](http://mpcm-evolution.org/OPM/Chapter11_OPM/data.zip)

```{r getdat}
if (!file.exists("data/phylo.nex")) {
    dir.create("data")
    download.file("http://mpcm-evolution.org/OPM/Chapter11_OPM/data.zip",
                  dest="data/OPM_ch11_data.zip")
    setwd("data")
    untar("OPM_ch11_data.zip")
    setwd("..")
}
phylo <- read.nexus("data/phylo.nex")
dat <- read.table("data/data_pois.txt",header=TRUE)
```

Add an observation-level random effect (not a bad idea in any case, but
in particular we want to match with `MCMCglmm` results).

```{r obsran}
dat$obs <- factor(seq(nrow(dat)))
```

Compute appropriate $Z$ matrix up front, to measure speed (also
reusable in a few places below):

```{r calc_phyloZ,cache=TRUE}
system.time(phyloZ <- phylo.to.Z(phylo))
```

### lme4 fit

```{r phylofit}
phylo_glmm_fit <- phylo_glmm(phen_pois~cofactor+(1|phylo)+(1|obs),
                             data=dat,family=poisson,phylo=phylo,
                             phyloZ=phyloZ)
```

```{r phylores}
lme4_res <- rbind(coef(summary(phylo_glmm_fit))[,1:2],
      cbind(matrix(unlist(VarCorr(phylo_glmm_fit))),NA))
```

From Garamszegi ch. 11 code examples. 
```{r MCMCglmm,cache=TRUE}
nitt <- 5e3 ## was 5e6
inv.phylo <- inverseA(phylo,nodes="TIPS",scale=TRUE)
prior <- list(G=list(G1=list(V=1,nu=0.02)),R=list(V=1,nu=0.02))
MCMC_time <- system.time(
    MCMCglmm_fit <- MCMCglmm(phen_pois~cofactor,random=~phylo,
                       family="poisson",ginverse=list(phylo=inv.phylo$Ainv),
                       prior=prior,data=dat,nitt=nitt,burnin=1000,
                       thin=nitt/1000,verbose=FALSE))
```
```{r MCMCres}
ss <- summary(MCMCglmm_fit)
MCMCglmm_res <- with(ss,rbind(solutions[,1:3],Gcovariances[,1:3],Rcovariances[,1:3]))
```

`MCMCglmm` fit takes about `r round(MCMC_time["elapsed"]/60)` minutes ... effective sample size is 1000, which means the initial suggested 5 million steps may be overkill ? (However, with only 5e4 parameter sets we get wonky-looking trace plots/effective sample size of only 200, so 5e5 may be necessary ...)

```{r glmmPQL,cache=TRUE}
rownames(dat) <- as.character(dat$phylo)
dat$allGrp <- factor(1) ## dummy grouping var because glmmPQL needs a group ...
glmmPQL_fit <- glmmPQL(phen_pois~cofactor,random=~1|phylo,
        data=dat,
        family="poisson",
        correlation=corBrownian(phy=phylo),
        verbose=FALSE)
```
```{r glmmPQL_res}
fixef(glmmPQL_fit)
```
Don't know if I got this right or not, but the answers look quite different ...  Use `gls()` with `varPower()` to
simulate Poisson errors?

```{r tmbfit,cache=TRUE}

### Don't need this in BB's update
fit_TMBstruc <- function(TMBStruc,verbose=FALSE) {
    obj <- with(TMBStruc, TMB:::MakeADFun(data.tmb, parameters, map = mapArg, 
                                          random = randomArg, profile = NULL, silent = !verbose, 
                                          DLL = "glmmTMB"))
    optTime <- system.time(fit <- with(obj, nlminb(start = par, 
                                                   objective = fn, gradient = gr)))
    fit$parfull <- obj$env$last.par.best
    sdr <- TMB:::sdreport(obj)
    return(list(fit=fit,sdr=sdr))
}

###

TMBstruc <- glmmTMB(phen_pois~cofactor+(1|phylo)+(1|obs),
                    data=dat,family=poisson,
                    debug=TRUE) # doFit=FALSE) in BB's update

TMBstruc_new <- modify_TMBstruc(TMBstruc,phylo,"phylo",phyloZ)

# glmmTMB_fit <- glmmTMB:::fitTMB(TMBstruc_new)
# tt <- tidy(glmmTMB_fit,scales=c(ran_pars="vcov",fixed=NA))
# glmmTMB_res <- tt[,c("term","estimate","std.error")]

glmmTMB_fit <- fit_TMBstruc(TMBstruc_new)
glmmTMB_res <- matrix(c(glmmTMB_fit$fit$par[1:2],exp(glmmTMB_fit$fit$par[3:4])^2),
                      ncol=1,
                      dimnames=list(c("intercept","cofactor","var.phylo","var.obs"),
                                    NULL))
```

## result comparison

```{r results}
glmmTMB_res
lme4_res
MCMCglmm_res
```

`lme4` and `glmmTMB` get very close to the same answer.
Close but not identical to `MCMCglmm`, in part because of priors ...


## Result comparison with Gaussian example in chapter 11

```{r G_results}
datG <- read.table("data/data_simple.txt",header=TRUE)
datG$obs <- factor(seq(nrow(datG)))
phylo_lmm_fit <- phylo_lmm(phen~cofactor+(1|phylo) +(1|obs),
                           data=datG,phylo=phylo,
                           control=lmerControl(check.nobs.vs.nlev="ignore",check.nobs.vs.nRE="ignore"),
                           phyloZ=phyloZ)
lme4_resLMM <- rbind(coef(summary(phylo_lmm_fit))[,1:2],
                     cbind(matrix(unlist(VarCorr(phylo_lmm_fit))),NA))

TMBstruc <- glmmTMB(phen~cofactor+(1|phylo)+(1|obs),
              data=datG,family=gaussian,
              debug=TRUE)
TMBstruc_new <- modify_TMBstruc(TMBstruc,phylo,"phylo",phyloZ)
glmmTMB_fit <- fit_TMBstruc(TMBstruc_new)
glmmTMB_res <- matrix(c(glmmTMB_fit$fit$par[1:2],exp(glmmTMB_fit$fit$par[3:4])^2),
                    ncol=1,
                    dimnames=list(c("intercept","cofactor","var.phylo","var.obs"),
                      NULL))

print(lme4_resLMM)
print(glmmTMB_res)

```


## To do

- Think about connections to/correspondence with Pagel's $\lambda$, which multiplies the phylogenetic covariance matrix by a factor $0<\lambda<1$?
- In an LMM what assumptions do we have to make about residual var, i.e. fix it to a small value?
- When is residual var unidentifiable (e.g. all terminal branch lengths identical)?
- how easily could we implement an O-U process (this would require that Z be recomputed each time with changing $\alpha$ [and reference level]: could be considerably more difficult/mess up some of the linear algebra tricks?)
- holes
     - `tidy`:
	      - method for `MCMCglmm_fit` should return SDs for random effects too
		  - clean up/document `scales` argument
     - `lme4` + `profile`?
- known/existing methods for PGLMM in R:
   - `pez::pglmm`
   - `glmmPQL/gls` + `ape::cor.*`
   - `MCMCglmm`
   - `spAMM::HLCor` ?
   - `lme4ord` (via Pagel's $\lambda$)
   - `ape::binaryPGLMM`


```{r check_retrms,echo=FALSE}
## utility function: what do we know that ought to be consistent?
check_retrms <- function(rt) {
    with(rt,{
        n.theta <- length(theta)
        n.re
        stopifnot(sum(sapply(Ztlist,nrow))==nrow(Zt))
        stopifnot(length(lower=n.theta))
        
    })
    return(NULL)
}
```
