---
title: "Phylogenetic comparative methods in the lme4-verse Part 2"
author: Michael Li 
date: "`r format(Sys.time(), '%H:%M %d %B %Y')`"
output:
    html_document:
    mathjax: default
---

## Setup

```{r pkgs,message=FALSE,warning=FALSE}
library(ape)
library(lme4)
library(Matrix)
## additional possibilities for PGLMMs ...
library(MCMCglmm)
library(MASS) ## glmmPQL + ape::corBrownian()
library(pez)
## n.b. at present need 'modular' devel branch:
## devtools::install_github("glmmTMB/glmmTMB/glmmTMB",ref="modular")
library(glmmTMB)
## utils
library(dplyr)
library(coda)
library(lattice)
library(broom) ## need install_github("bbolker/broom")
library(dotwhisker)
library(nlme)
```

```{r phylocode,echo=FALSE,warning=FALSE,message=FALSE}
phylo.to.Z <- function(r) {
    ntip <- length(r$tip.label)
    Zid <- Matrix(0.0,ncol=length(r$edge.length),nrow=ntip)
    nodes <- (ntip+1):max(r$edge)
    root <- nodes[!(nodes %in% r$edge[,2])]
    for (i in 1:ntip) {
        cn <- i  ## current node
        while (cn != root) {
            ce <- which(r$edge[,2]==cn)   ## find current edge
            Zid[i,ce] <- 1   ## set Zid to 1
            cn <- r$edge[ce,1]            ## find previous node
        }
    }
    Z <- t(r$edge.length * t(Zid))
    return(Z)
}

#' split a square (block) matrix into component blocks 
#' @param M square matrix
#' @param ind indices (0,n1,n2,...) giving the endpoint of each block
split_blkMat <- function(M,ind) {
    res <- list()
    if (length(ind)==1) return(list(M))
    for (i in 1:(length(ind)-1)) {
        v <- (ind[i]+1):ind[i+1]
        res[[i]] <- M[v,v]
    }
    return(res)
}

#' modify reTrms object
#' @param rt a reTrms object
#' @param phylo a phylo object (phylogenetic tree)
#' @param phylonm name of phylogenetic term in model
#' @param phyloZ Z matrix built on branch length
modify_phylo_retrms <- function(rt,phylo,phylonm,
                                phyloZ=phylo.to.Z(phylo),sp) {
    rep_phylo <- rt$Zt@Dim[2]/length(unique(sp)) ## number of column (aka same as number of obs)
    ## FIXME: better way to specify phylonm
    ## need to replace Zt, Lind, Gp, flist, Ztlist
    ## we have the same number of parameters (theta, lower),
    ##  same number of obs
    n.edge <- nrow(phylo$edge)
    phylo.pos <- which(names(rt$cnms)==phylonm)
    inds <- c(0,cumsum(sapply(rt$Ztlist,nrow)))
    ## Zt: substitute phylo Z for previous dummy (scalar-intercept) Z
    # for(i in phylo.pos){
    # repterms <- nrow(rt[["Ztlist"]][[i]])/length(unique(sp))
    # rt[["Ztlist"]][[i]] <- KhatriRao(do.call(cbind,replicate(rep_phylo,t(phyloZ),simplify = FALSE)),
    #             matrix(1
    #                    , ncol=ncol(rt[["Ztlist"]][[i]])
    #                    , nrow=repterms)
    # )
    ## reconstitute Zt from new Ztlist
    # }
    # rt[["Zt"]] <- do.call(rbind,rt[["Ztlist"]])
    ## Gp: substitute new # random effects (n.edge) for old # (n.phylo)
    Gpdiff <- diff(rt$Gp)  ## old numbers
    Gpdiff_new <- Gpdiff
    # for(i in phylo.pos){
    # Gpdiff_new[i] <- n.edge  ## replace
    # }
    rt[["Gp"]] <- as.integer(c(0,cumsum(Gpdiff_new)))          ## reconstitute
    ## Lind: replace phylo block with the same element, just more values
    Lind_list <- split(rt[["Lind"]],rep(seq_along(Gpdiff),Gpdiff))
    # for(i in phylo.pos){
    # Lind_list[[i]] <- rep(Lind_list[[i]][1],n.edge)
    # }
    # rt[["Lind"]] <- unlist(Lind_list)
    ## Lambdat: replace block-diagonal element in Lambdat with a
    ##   larger diagonal matrix
    Lambdat_list <- split_blkMat(rt[["Lambdat"]],inds)
    # for(i in phylo.pos){
    # Lambdat_list[[i]] <- KhatriRao(Diagonal(n.edge,1.0),
    #             matrix(1
    #                    , ncol=n.edge
    #                    , nrow=repterms))
    # }
    
    for(i in phylo.pos){
    repterms <- nrow(rt[["Ztlist"]][[i]])/length(unique(sp))
#     rt[["Ztlist"]][[i]] <- KhatriRao(do.call(cbind,replicate(rep_phylo,t(phyloZ),simplify = FALSE)),
#                 matrix(1
#                        , ncol=ncol(rt[["Ztlist"]][[i]])
#                        , nrow=repterms)
#     )
    ## reconstitute Zt from new Ztlist
    rt[["Ztlist"]][[i]] <- t(KhatriRao(phyloZ,matrix(1,ncol=ncol(phyloZ),nrow=repterms))) %*% rt[["Ztlist"]][[i]]
    Gpdiff_new[i] <- n.edge  ## replace
    Lind_list[[i]] <- rep(Lind_list[[i]][1],n.edge)
    Lambdat_list[[i]] <- KhatriRao(diag(n.edge),
                Matrix(1
                       , ncol=n.edge
                       , nrow=repterms))
    }
    rt[["Zt"]] <- do.call(rbind,rt[["Ztlist"]])
    rt[["Lind"]] <- unlist(Lind_list)
    rt[["Lambdat"]] <- Matrix::.bdiag(Lambdat_list)
    ## flist: 
    rt[["flist"]] <- as.list(rt[["flist"]])
    rt[["flist"]][[phylonm]] <- factor(paste0("edge_",seq(n.edge)))
    return(rt)
}

#' 
phylo_glmm <- function(formula,data,family,phylo,phylonm,phyloZ,sp) {
    glmod <- glFormula(formula=formula,data = data, family = family)
    glmod$reTrms <- modify_phylo_retrms(glmod$reTrms,phylo,
                                        phylonm,phyloZ,sp)
    devfun <- do.call(mkGlmerDevfun, glmod)
    opt <- optimizeGlmer(devfun)
    devfun <- updateGlmerDevfun(devfun, glmod$reTrms)
    opt <- optimizeGlmer(devfun, stage=2)
    mkMerMod(environment(devfun), opt, glmod$reTrms, fr = glmod$fr)
}

phylo_lmm <- function(formula,data,phylo,phylonm,phyloZ,control,sp) {
  lmod <- lFormula(formula=formula,data = data,control=control)
  lmod$reTrms <- modify_phylo_retrms(lmod$reTrms,phylo,
                                     phylonm,phyloZ,sp)
  devfun <- do.call(mkLmerDevfun, lmod)
  opt <- optimizeLmer(devfun)
  # devfun <- updateLmerDevfun(devfun, lmod$reTrms)
  # opt <- optimizeLmer(devfun, stage=2)
  mkMerMod(environment(devfun), opt, lmod$reTrms, fr = lmod$fr)
}
```

```{r getdat}
if (!file.exists("data/phylo.nex")) {
    dir.create("data")
    download.file("http://mpcm-evolution.org/OPM/Chapter11_OPM/data.zip",
                  dest="data/OPM_ch11_data.zip")
    setwd("data")
    untar("OPM_ch11_data.zip")
    setwd("..")
}
phylo <- read.nexus("data/phylo.nex")
dat_pois <- read.table("data/data_pois.txt",header=TRUE)
dat_pois$obs <- factor(seq(nrow(dat_pois)))
dat_gaus <- read.table("data/data_simple.txt",header=TRUE)
dat_gaus$obs <- factor(seq(nrow(dat_gaus)))
```

```{r calc_phyloZ}
system.time(phyloZ <- phylo.to.Z(phylo))
```

## Poisson Example
### lme4 fit

```{r phylofit_pois}
basic_pois_fit <- glmer(phen_pois~cofactor+(1|phylo),
                        data=dat_pois,family=poisson)

basic_pois_res <- rbind(coef(summary(basic_pois_fit))[,1:2],
      cbind(matrix(unlist(VarCorr(basic_pois_fit))),NA))

phylo_pois_fit <- phylo_glmm(phen_pois~cofactor+(1|phylo)+(1|obs),sp=dat_pois$phylo,phylonm="phylo",
                             data=dat_pois,phylo=phylo,family=poisson,
                             phyloZ=phyloZ)

lme4_pois_res <- rbind(coef(summary(phylo_pois_fit))[,1:2],
      cbind(matrix(unlist(VarCorr(phylo_pois_fit))),NA))

print(basic_pois_res)
print(lme4_pois_res)
```
### MCMCglmm

```{r MCMCglmm_posi,cache=TRUE}
nitt <- 1e5 ## was 5e6
inv.phylo <- inverseA(phylo,nodes="TIPS",scale=TRUE)
prior <- list(G=list(G1=list(V=1,nu=0.02)),R=list(V=1,nu=0.02))
MCMC_time <- system.time(
    MCMCglmm_fit_pois <- MCMCglmm(phen_pois~cofactor,random=~phylo,
                       family="poisson",ginverse=list(phylo=inv.phylo$Ainv),
                       prior=prior,data=dat_pois,nitt=nitt,burnin=1000,
                       thin=nitt/1000,verbose=FALSE))

ss <- summary(MCMCglmm_fit_pois)
MCMCglmm_pois_res <- with(ss,rbind(solutions[,1:3],Gcovariances[,1:3],Rcovariances[,1:3]))

print(MCMCglmm_pois_res)
```
### glmmPQL

```{r glmmPQL_pois,cache=TRUE}
rownames(dat_pois) <- as.character(dat_pois$phylo)
dat_pois$allGrp <- factor(1) ## dummy grouping var because glmmPQL needs a group ...
glmmPQL_poisnocor_fit <- glmmPQL(phen_pois~cofactor,random=~1|phylo,
        data=dat_pois,
        family="poisson",
        # correlation=corBrownian(phy=phylo),
        verbose=FALSE)

glmmPQL_poiscor_fit <- glmmPQL(phen_pois~cofactor,random=~1|phylo,
        data=dat_pois,
        family="poisson",
        correlation=corBrownian(phy=phylo),
        verbose=FALSE)

print(fixef(glmmPQL_poisnocor_fit))
print(fixef(glmmPQL_poiscor_fit))
```

## Gaussian Example

### lme4 fit

```{r phylofit_gaus}
basic_gaus_fit <- lmer(phen~cofactor+(1|phylo),
                        data=dat_gaus,
control=lmerControl(check.nobs.vs.nlev="ignore",check.nobs.vs.nRE="ignore"))

basic_gaus_res <- rbind(coef(summary(basic_gaus_fit))[,1:2],
      cbind(matrix(unlist(VarCorr(basic_gaus_fit))),NA))

phylo_gaus_fit <- phylo_lmm(phen~cofactor+(1|phylo)+(1|obs),sp=dat_gaus$phylo,
                             data=dat_gaus,phylo=phylo,phylonm="phylo",
control=lmerControl(check.nobs.vs.nlev="ignore",check.nobs.vs.nRE="ignore"),
                           phyloZ=phyloZ)

lme4_gaus_res <- rbind(coef(summary(phylo_gaus_fit))[,1:2],
      cbind(matrix(unlist(VarCorr(phylo_gaus_fit))),NA))

print(basic_gaus_res)
print(lme4_gaus_res)
```
### MCMCglmm

```{r MCMCglmm_gaus,cache=TRUE}
nitt <- 1e5 ## was 5e6
inv.phylo <- inverseA(phylo,nodes="TIPS",scale=TRUE)
prior <- list(G=list(G1=list(V=1,nu=0.02)),R=list(V=1,nu=0.02))
MCMC_time <- system.time(
    MCMCglmm_fit_gaus <- MCMCglmm(phen~cofactor,random=~phylo,
                       family="gaussian",ginverse=list(phylo=inv.phylo$Ainv),
                       prior=prior,data=dat_gaus,nitt=nitt,burnin=1000,
                       thin=nitt/1000,verbose=FALSE))

ss <- summary(MCMCglmm_fit_gaus)
MCMCglmm_gaus_res <- with(ss,rbind(solutions[,1:3],Gcovariances[,1:3],Rcovariances[,1:3]))

print(MCMCglmm_gaus_res)
```
### glmmPQL

```{r glmmPQL_gaus,cache=TRUE}
rownames(dat_gaus) <- as.character(dat_gaus$phylo)
dat_gaus$allGrp <- factor(1) ## dummy grouping var because glmmPQL needs a group ...
glmmPQL_gausnocor_fit <- glmmPQL(phen~cofactor,random=~1|phylo,
        data=dat_gaus,
        family="gaussian",
        # correlation=corBrownian(phy=phylo),
        verbose=FALSE)

glmmPQL_gauscor_fit <- glmmPQL(phen~cofactor,random=~1|phylo,
        data=dat_gaus,
        family="gaussian",
        correlation=corBrownian(phy=phylo),
        verbose=FALSE)

print(fixef(glmmPQL_gausnocor_fit))
print(fixef(glmmPQL_gauscor_fit))
```
### gls

```{r gls_gaus}
gls_gausnocor_fit <- gls(phen~cofactor,
        data=dat_gaus,
        # correlation=corBrownian(phy=phylo),
        verbose=FALSE)

gls_gauscor_fit <- gls(phen~cofactor,
        data=dat_gaus,
        correlation=corBrownian(phy=phylo),
        verbose=FALSE)

print(summary(gls_gausnocor_fit))
print(summary(gls_gauscor_fit))


```

## Results
### poisson results

```{r poisson_res}
tt_glmer_nocor_pois <- tidy(basic_pois_fit)
tt_glmer_cor_pois <- tidy(phylo_pois_fit)
tt_MCMCglmm_cor_pois <- tidy(MCMCglmm_fit_pois)

print(tt_glmer_nocor_pois)
print(tt_glmer_cor_pois)
print(tt_MCMCglmm_cor_pois)

tt_comb <- bind_rows(glmer_nocor=tt_glmer_nocor_pois,
                     glmer_cor=tt_glmer_cor_pois,
                     MCMCglmm_cor=tt_MCMCglmm_cor_pois,
                     .id="model") %>% filter(effect=="fixed")

print(dwplot(tt_comb)+ theme_bw())
```
### Gaussian results

```{r gaus_res}
tt_lmer_nocor_gaus <- tidy(basic_gaus_fit)
tt_lmer_cor_gaus <- tidy(phylo_gaus_fit)
tt_MCMCglmm_cor_gaus <- tidy(MCMCglmm_fit_gaus)

print(tt_lmer_nocor_gaus)
print(tt_lmer_cor_gaus)
print(tt_MCMCglmm_cor_gaus)

tt_comb <- bind_rows(lmer_nocor=tt_lmer_nocor_gaus,
                     lmer_cor=tt_lmer_cor_gaus,
                     MCMCglmm_cor=tt_MCMCglmm_cor_gaus,
                     .id="model") %>% filter(effect=="fixed")

print(dwplot(tt_comb)+ theme_bw())
```


## Pez example

```{r pez,echo=FALSE,warning=FALSE,message=FALSE,results="hide",include=FALSE}

example("communityPGLMM",echo = FALSE)

```

```{r pez test,echo=FALSE,warning=FALSE,message=FALSE}

pez_gausREsp <- communityPGLMM(Y ~ X, data = dat, family = "gaussian", sp = dat$sp,
                       site = dat$site, random.effects = list(re.1, re.3), REML = FALSE,
                       verbose = FALSE)
print(summary(pez_gausREsp))
lmer_gausREsp <- lmer(Y ~ X + (1 | sp) + (0 + X | sp), data=dat, REML = FALSE)
print(summary(lmer_gausREsp))

pez_gausREspint <- communityPGLMM(Y ~ X, data = dat, family = "gaussian", sp = dat$sp,
                       site = dat$site, random.effects = list(re.1), REML = FALSE,
                       verbose = FALSE)
print(summary(pez_gausREspint))
lmer_gausREspint <- lmer(Y ~ X + (1 | sp), data=dat, REML = FALSE)
print(summary(lmer_gausREspint))

```


```{r rep sp matrix,echo=FALSE, warning=FALSE,message=FALSE}
## REsp (1|sp) + (0+X|sp)
## dim Zt = 30x150
## Ztlist,1 = 1|sp 15x150
## Ztlist,2 = 0 + X | sp 15x150

phyZ <- phylo.to.Z(phy)

dat$obs <- dat$sp
pez_gausREspintcor <- communityPGLMM(Y ~ X, data = dat, family = "gaussian", sp = dat$sp,
                       site = dat$site, random.effects = list(re.1,re.2,re.3,re.4), REML = FALSE,
                       verbose = FALSE)
print(summary(pez_gausREspintcor))
lmer_gausREspintcor <- phylo_lmm(Y ~ X + (1+X|sp) + (1+X|obs)
                                 , data=dat
                                 , phylonm = "sp"
                                 , sp = dat$sp
                                 , phylo = phy
                                 , phyloZ=phyZ
                                 , control=lmerControl(check.nobs.vs.nlev="ignore",check.nobs.vs.nRE="ignore"))
print(summary(lmer_gausREspintcor))

```

## Todo
- show the estimates with/without phylo-correlation structure
- need to add TMB (ATM TMB and lme4 results are the same)
- no tidy method for glmmPQL and gls 
- cannot do random slopes (done)
- cannot do more than one random effect, for example: 1|speices + 1|site (done)
- binomial (glmer/ phylo_glmm) get different fits compare to pez (communityPGLMM)
