---
title: "Wald errors of variances"
author: "Ben Bolker"
date: "`r format(Sys.time(), '%H:%M %d %B %Y')`"
output: html_document
---

## Introduction

`lme4` users frequently request the capability to compute the variances (standard errors, variance-covariance matrix) of the random-effects variance parameters. `lme4` has switched to providing profile and parametric bootstrap, but *not* Wald confidence intervals on the random-effect variances, because the Wald intervals on variance-covariance parameters are often really bad. Nevertheless, one sometimes does want the asymptotic (Wald) variance-covariance matrix of the variance-covariance parameters ...  This document shows how to do that for `merMod` objects, and discusses the issues more generally.

In the `nlme` package the Wald variances are (implicitly) available via the `intervals()` function in the `nlme` package, which returns confidence intervals based on Wald intervals on the transformed ("unconstrained" scale, i.e. the log scale for variance parameters and the (tanh(x)/2) scale for correlation parameters; the underlying approximate variance-covariance matrices (on the unconstrained scale) are stored as the `$apVar` component in the model object.

The basic problem is that `lme4`'s deviance function uses a scaled Cholesky-factor parameterization, e.g. for a random slope model the variance-covariance matrix is

$$
\Sigma = \sigma^2 L L^T = \sigma^2 
\left( 
\begin{array}{cc} \theta_1 & 0  \\ \theta_2 & \theta_3  \end{array} 
\right)
\left(
\begin{array}{cc} \theta_1 & \theta_2  \\ 0 & \theta_3 \end{array}
\right)
$$

It's easy enough to get an approximate Hessian of the $\theta$ parameters by finite differences, but not so easy to get estimates of the variance-covariance parameters.  There are a few strategies for converting from this to the variance scale:

* Use an internal deviance function (`devfun2`) that is parameterized on the $\sigma$ scale.  (At present `devfun2` [only works for ML, not REML](https://github.com/lme4/lme4/issues/325); this could be hacked ...)
* Use the delta method on the equation above (but we need the variance in the $\sigma^2$ parameter ... ?)
* Compute dependence of the $\sigma$-scale parameters on the $\theta$ parameters and use the delta method.

## Preliminaries

```{r pkgs,message=FALSE}
library("lme4")
library("numDeriv")
library("ggplot2"); theme_set(theme_bw())
```

Fit the good old `sleepstudy` model:
```{r fitmodels,message=FALSE}
fm1 <- lmer(Reaction ~ Days + (Days | Subject), sleepstudy)
```

For comparison with the `devfun2` likelihood profile (which uses ML rather than REML) we also fit the ML equivalent:
```{r mlfit}
fm1.ML <- refitML(fm1)
```

Also fit with `lme` so that we can compare the results of `intervals()`:
```{r lmefit,message=FALSE}
library(nlme)
fm1.lme <- lme(Reaction ~ Days, random = ~Days| Subject, sleepstudy)
fm1.lme.ML <- update(fm1.lme,method="ML")
``` 

## Delta method

Suppose we have parameters $\mathbf \theta$ (possibly also including fixed-effect parameters in this case; I mean the full set of parameters that determine the deviance), and a variance-covariance matrix $\Sigma_{\mathbf \theta}$ for them, but we have a set of derived parameters $\mathbf \sigma$ and we want to get $\Sigma_{\mathbf \sigma}$.  Then all we actually need is the Jacobian matrix ${\mathbf J} = ( \partial \sigma_i/\partial \theta_j )$, which we can get by finite differencing; then $\mathbf J^T \Sigma_{\mathbf \theta} \mathbf J$ should give us  $\Sigma_{\mathbf \sigma}$.

We can also get the derivatives of $\sigma_{ij}$ with respect to $\theta$ analytically (see below).
Unfortunately, for a LMM the actual variance-covariance parameters are scaled by the residual variance $\sigma^2$;
if we're willing to ignore the uncertainty in $\sigma^2$ we can simply scale the Jacobian by $\sigma$ ...

$$
\left(
\begin{array}{ccc}
\sigma^2_x & \sigma_{xy} & \sigma_{xz} \\
\sigma_{xy} & \sigma^2_y & \sigma_{yz} \\
\sigma_{xz} & \sigma_{yz} & \sigma^2_z 
\end{array}
\right)
=
\left(
\begin{array}{ccc}
\theta_1 & 0 & 0 \\
\theta_2 & \theta_4 & 0 \\
\theta_3 & \theta_5 & \theta_6 
\end{array}
\right)
\left(
\begin{array}{ccc}
\theta_1 & \theta_2 & \theta_3 \\
0 & \theta_4 & \theta_5 \\
0 & 0 & \theta_6
\end{array} 
\right)
\qquad ,
$$

What are the derivatives of $\{\sigma^2_x, \sigma_{xy}, \sigma_{xz}, \ldots\}$
with respect to $\{\theta_i\}$?

It turns out this has a nice sparsity structure, made up of lower-triangular blocks
based on each of the original parameters.
For the $3 \times 3$ case (with rows=$\theta$ values, columns=$\sigma_{\cdot \cdot}$ values
(lower triangle, column-wise order), this is
$$
\left(
\begin{array}{cccccc}
2 \theta_1 & 0 & 0 & 0 & 0 & 0 \\
\theta_1   & \theta_1 & 0 & 0 & 0 & 0 \\
\theta_1   & 0 & \theta_1 & 0 & 0 & 0 \\
0 & 2 \theta_2 & 0 & 2 \theta_4 & 0 & 0 \\
0 & \theta_2 & \theta_2 & \theta_4 & \theta_4 & 0 \\
0 & 0 & 2 \theta_3 & 0 & 2 \theta_5 & 2 \theta_6
\end{array}
\right)
$$

```{r cholblock,echo=FALSE}
bmat <- function(t,n) {
    m <- matrix(0,n,n)
    diag(m) <- t
    m[,1] <- t
    m[1,1] <- 2*t
    return(m)
}
bigmat <- function(t) {
    n <- (sqrt(8*length(t)+1)-1)/2
    N <- n*(n+1)/2
    M <- matrix(0,N,N)
    svec <- n:1
    posvec <- cumsum(c(1,n:2))
    tc <- 1
    for (i in 1:n) {  ## horizontal blocks
        ipos <- posvec[i]
        for (j in i:n) { ## vertical blocks
            jpos <- posvec[j]
            hadj <- (j-i)
            yy <- (ipos+hadj):(ipos+hadj+svec[j]-1)
            xx <- jpos:(jpos+svec[j]-1)
            M[xx,yy] <- bmat(t[tc],svec[j])
            tc <- tc+1
        }
    }
    return(M)
}
```

In particular:

```{r update}
## generate a model with a modified set of theta parameters
update.lmerMod <- function(object,theta,...) {
    if (missing(theta)) return(update.default(object,...))
    object2 <- object
    ## deep-copy the (only) reference-class slots ...
    object2@pp <- object@pp$copy()
    object2@resp <- object@resp$copy()
    object2@pp$setTheta(theta)
    dd <- as.function(object2)
    dval <- dd(theta)  ## update internal structures
    ## object2@resp$updateMu(object2@resp$mu)  ## ?? not helping/not necessary
    
    mkMerMod(environment(dd),
             opt=list(par=theta,fval=dval,conv=0),
             fr=model.frame(object2),
             reTrms=getME(object2,
                          c("Zt","Lambdat","Lind","theta",
                            "lower","flist","cnms","Gp"))
    )
}
## for GLMMs: update.glmerMod will need a beta parameter;
## object@pp$setBeta0(beta)
## dd <- update(object,devFunOnly=TRUE)
## parvec <- c(theta,beta)
## dval <- dd(parvec)
## opt=list(par=parvec,...)

## test: is it safe?
fm2 <- update(fm1,theta=c(2,2,2))
orig_theta <- getME(fm1,"theta")
stopifnot(identical(unname(getME(fm2,"theta")),c(2,2,2)))
stopifnot(identical(orig_theta,getME(fm1,"theta")))
tn <- function(object) {
    c(names(getME(object,"theta")),"sigma")
}
## compute standard deviations/correlations for a given theta vector
outfun <- function(object,t) {
    newmod <- update(object,theta=t)
    av <- as.data.frame(VarCorr(newmod),
                        order="lower.tri")
    r <- setNames(av[,"sdcor"],tn(object))
    return(r)
}
```

Test derivative calculations for $\sigma$:
```{r test_outfun,cache=TRUE,fig.width=10,fig.height=3}
pp <- getME(fm1.ML,"theta")
Jmat <- jacobian(outfun,pp,object=fm1.ML)
slicefun <- function(pos) {
    tvec <- seq(0.8*pp[pos],1.2*pp[pos],length=51)
    svec <- sapply(tvec,function(x) {
                       pp2 <- pp
                       pp2[pos] <- x
                       outfun(fm1.ML,t=pp2)["sigma"]
                       })
    list(x=tvec,y=svec)
}
ss <- lapply(1:3,slicefun)
par(las=1,bty="l",mfrow=c(1,3))
for (i in 1:3) {
    plot(ss[[i]],xlab=paste0("theta[",i,"]"),ylab="sigma")
    curve((x-pp[i])*Jmat[4,i]+sigma(fm1.ML),add=TRUE,col=2)
}
```

Does my `update` function correctly update sigma?
Apparently so ...

```{r test_update}
up1 <- update(fm1,theta=c(2,2,2))
## update 'from scratch'
up2 <- suppressWarnings(update(fm1,
                               start=c(2,2,2),
                               control=lmerControl(optCtrl=list(maxfun=1))))
identical(unname(getME(up2,"theta")),c(2,2,2))
all.equal(sigma(up1),sigma(up2))
```

```{r waldvar1}
waldVar1 <- function(object) {
    pp <- getME(object,"theta") ## GLMM: unlist(getME(dd,c("theta","beta")))
    Jmat <- jacobian(outfun,pp,object=object)
    dimnames(Jmat) <- list(tn(object),paste0("theta",seq(ncol(Jmat))))
    dd <- as.function(object) ## deviance/REMLcrit function
    hh <- 1/2*hessian(dd,pp)  ## ... calculate information matrix ...
                              ## 1/2 = deviance to log-lik scale)
    ## vv <- solve(hh)        ## invert to get Wald vars of theta parameters
    ## m2 <- Jmat %*% vv %*% t(Jmat)  ## delta method
    ## slightly better linear algebra:
    m1 <- Jmat %*% solve(hh,t(Jmat))
    return(m1)
}
wsd1 <- sqrt(diag(waldVar1(fm1)))
wsd1.ML <- sqrt(diag(waldVar1(fm1.ML)))
```
As we'll see, the estimate of the standard deviation for `sigma` 
is off by a factor of about 2 (but not exactly) compared to both the second method
(using `devfun2`) and the `lme` results.
I can't figure out why right now - this would occur naturally if
I got confused between variance and standard deviation scale,
but (a) it should be an exact factor of 2 (b) if so I can't find where it's happening.

### Finite differences on the $\sigma$ scale

The other way to do this is using the internal `devfun2()` function, which is used within `profile` --- it is essentially the same as the basic `devfun()`, but it uses a standard deviation/correlation parameterization (and adds the residual standard deviation as a parameter) rather than a Cholesky parameterization:
```{r devfun2}
waldVar2 <- function(object) {
    ## test for/warn if ML fit?
    dd <- lme4:::devfun2(object,useSc=TRUE,signames=FALSE)
    nvp <- length(attr(dd,"thopt"))+1 ## variance parameters (+1 for sigma)
    pars <- attr(dd,"optimum")[seq(nvp)] ## var params come first
    hh <- hessian(dd,pars)
    ## factor of 2: deviance -> negative log-likelihood
    vv <- 2*solve(hh)
    nn <- tn(object)
    dimnames(vv) <- list(nn,nn)
    return(vv)
}
## would be identical applied to fm1 (would get ML-ized anyway)
wsd2.ML <- sqrt(diag(waldVar2(fm1.ML)))
```

At the moment `devfun2()` cannot be used for REML fits
(as it is designed for computing likelihood profiles, that wouldn't
make much sense).

### Convert scales

I have a bunch of code now (not shown here) for converting
estimates+standard deviations to confidence intervals on 
a different scale. We translate the estimates and standard
deviations to the link scale; compute the Wald confidence
intervals on that scale; and back-transform.

```{r linkconv,echo=FALSE}
loglink <- poisson()[c("linkfun","linkinv","mu.eta")]
tanhlink <- list(linkfun=function(x) tanh(x/2),
                 linkinv=function(x) 2*atanh(x),
                 mu.eta=function(x) -2/(x^2-1) ## 1/(cosh(x)+1))
                 )
sqrtlink <- list(linkfun=sqrt,
                linkinv=function(x) x^2,
                mu.eta=function(x) 2*x)
identitylink <- list(linkfun=identity,linkinv=identity,
                     mu.eta=function(x) 1)
linktest <- function(link,value,verbose=FALSE) {
    with(link,
         {
             a1 <- all.equal(linkfun(linkinv(value)),value)
             a2 <- all.equal(grad(linkinv,value),mu.eta(value))
             if (verbose) {
                 cat("link/inv-link: ")
                 print(a1)
                 cat("gradient: ")
                 print(a2)
             }
             isTRUE(a1) && isTRUE(a2)
         })
}
stopifnot(linktest(loglink,0.8))
stopifnot(linktest(tanhlink,0.8))
stopifnot(linktest(sqrtlink,0.8))

conv <- function(est,sd,link,inv=TRUE) {
    with(link, {
         if (inv) {
             sd <- sd*mu.eta(est)
             est <- linkinv(est)
         } else {
             sd <- sd/mu.eta(linkfun(est))
             est <- linkfun(est)
         }
         list(est=est,sd=sd)
     })
}
## round-trip:
c0 <- list(est=0.5,sd=0.2)
c1 <- do.call(conv,c(c0,list(link=loglink,inv=TRUE)))
c2 <- do.call(conv,c(c1,list(link=loglink,inv=FALSE)))
stopifnot(all.equal(c0,c2))
c1 <- do.call(conv,c(c0,list(link=tanhlink,inv=TRUE)))
c2 <- do.call(conv,c(c1,list(link=tanhlink,inv=FALSE)))
stopifnot(all.equal(c0,c2))

##' variation of confint.default that allows for transformation
##' @param object an object with \code{coef} and \code{vcov} methods
##' @param parm a vector of numeric or character labels for parameters
##' @param level
##' @param either a single list with \code{linkinv}, \code{linkfun},
##' and \code{mu.eta} components, or a list with elements \code{trans}
##' (a list of transformation objects) and \code{pars} (a list of
##' numeric vectors of parameter indices corresponding to each transformation)
confint2 <- function (object, parm, level = 0.95,
          trans = NULL, ...) 
{
    cf <- coef(object)
    pnames <- names(cf)
    if (missing(parm)) 
        parm <- pnames
    else if (is.numeric(parm)) 
        parm <- pnames[parm]
    a <- (1 - level)/2
    a <- c(a, 1 - a)
    pct <- stats:::format.perc(a, 3)
    fac <- qnorm(a)
    ci <- array(NA, dim = c(length(parm), 2L),
                dimnames = list(parm, pct))
    ses <- sqrt(diag(vcov(object)))[parm]
    ci[] <- cf[parm] + ses %o% fac
    if (!is.null(trans)) {
        if (all(c("linkinv","linkfun","mu.eta") %in% names(trans))) {
            ## trans is given just as a link object -- convert
            ## to full list (excluding param #1 by default)
            npar <- length(cf)
            trans <- list(links=trans,pars=2:npar)
        }
        for (i in seq_along(trans$pars)) {
            p <- trans$pars[[i]]
            lf <- trans$links[[i]]
            for (j in p) {
                cc <- conv(cf[j],ses[j],link=lf,inv=FALSE)
                ci[j,] <- lf$linkinv(cc$est + fac*cc$sd)
            }
        }
    }
    ci
}
## test
set.seed(101)
dd <- data.frame(x=1:10,y=rnorm(10))
m1 <- lm(y~x,dd)
stopifnot(all.equal(confint2(m1),confint.default(m1)))
```


### Results and comparisons

Now that we have the Wald variance-covariance matrix, I'm going
to cook up a little bit of structure that will let us use 
`confint.default` to compute the Wald intervals:

```{r confints}
pars <- as.data.frame(VarCorr(fm1.ML),order="lower.tri")[,"sdcor"]
names(pars) <- tn(fm1.ML)
vhack1 <- list(coefficients=pars,vcov=waldVar1(fm1.ML))
vhack2 <- list(coefficients=pars,vcov=waldVar2(fm1.ML))
vcov.default <- function(object,...) object$vcov
## convert variance parameters to log scale,
## correlation to atanh scale (to match lme)
trans_list1 <- list(links=list(loglink,tanhlink),
                    pars=list(c(1,3,4),2))
(wc1S <- confint2(vhack1,trans=trans_list1))
(wc2S <- confint2(vhack2,trans=trans_list1))
## non-scaled versions for comparison
wc1 <- confint2(vhack1)
wc2 <- confint2(vhack2)
```

```{r intervals}
(lmeint <- intervals(fm1.lme.ML,which="var-cov"))
```

#### Comparison with likelihood profile

Compute the likelihood profile:
```{r fitprof,cache=TRUE}
## automatically ML-ified
pp <- profile(fm1,which="theta_",signames=FALSE)
```

The linear approximation to the profile for a focal parameter $x$ should
be $\zeta \approx (x-\hat x)/\sigma_x$, where $\hat x$ is the MLE and
$\sigma_x$ is the Wald standard deviation:
```{r profplot}
dfprof <- as.data.frame(pp)
mframe <- data.frame(.par=levels(dfprof$.par),ctr=pars,slope=1/wsd2.ML)
dfprof <- merge(dfprof,mframe)
ggplot(dfprof,aes(.focal,.zeta))+geom_line()+geom_point(size=0.8)+
    facet_wrap(~.par,scale="free")+
    geom_hline(yintercept=0,color="gray")+
    geom_hline(yintercept=c(-1.96,1.96),color="gray",lty=2)+
    geom_line(aes(y=(.focal-ctr)*slope),color="red")
```    

In this particular example, there's not a huge divergence from
linearity over the range defined by the 95% confidence intervals
(dashed horizontal lines) ... as confirmed by the confidence intervals.
(Should check with `waldVar1`, sort out problem with `sigma` ...)

#### Comparison of CIs

```{r cmpconf,echo=FALSE,fig.height=3,fig.width=10}
pci <- confint(pp)
nf  <- function(x) setNames(as.data.frame(x),c("lwr","upr"))
lme4Dat <- rbind(
   data.frame(method="profile_lme4",
              par=tn(fm1),est=pars,nf(pci)),
    data.frame(method="Wald1",
               par=rownames(wc1S),est=pars,nf(wc1S)),
    data.frame(method="Wald2",
               par=rownames(wc2S),est=pars,nf(wc2S)),
    data.frame(method="Wald1_noscale",
               par=rownames(wc1),est=pars,nf(wc1)),
    data.frame(method="Wald2_noscale",
               par=rownames(wc2),est=pars,nf(wc2)))

lmeDat <- data.frame(method="Wald_lme",
     par=tn(fm1),est=pars, ## close enough to identical
     rbind(nf(lmeint$reStruct$Subject[c(1,3,2),c(1,3)]),
           nf(rbind(lmeint$sigma[c(1,3)]))))
allDat <- rbind(lme4Dat,lmeDat)
newlev <- c("Wald1_noscale","Wald1","Wald2_noscale","Wald2",
            "Wald_lme","profile_lme4")
allDat <- transform(allDat,method=factor(method,level=newlev))
ggplot(allDat,aes(par,est,ymin=lwr,ymax=upr,colour=method))+
    geom_pointrange(position=position_dodge(width=0.5))+
    facet_wrap(~par,scale="free",nrow=1)+
        scale_x_discrete(breaks=NULL)+
            labs(x="",y="Standard deviation/correlation")+
                scale_colour_brewer(palette="Paired")
```

Discrepancies:

* for this case, the profile CIs are actually very
close to the appropriately link-scaled CIs overall
* the discrepancy in the $\sigma$ CI from `waldVar1`
is obvious; it stems from the discrepancy in the 
estimated std dev of $\sigma$ (`r round(wsd1.ML["sigma"],2)`
for `waldVar1` vs. `r round(wsd2.ML["sigma"],2)` for
`waldVar2`)
* the largest other discrepancies are for
the correlation parameter.  They're close, but:
did I get the `lme` link scale correct?


### Continuing

(*Don't remember what I was doing this for: trying to get higher
derivatives to test for optimal transformations?*)
Second derivative via chain rule?  We're looking for $\frac{\partial^2 f(g(x))}{\partial x^2}$ (or $\frac{\partial^2 (f \circ g)(x)}{\partial x^2}$ may be easier notation in this case).

I should have been able to figure this out myself, but looking at
[Faà di Bruno's formula](http://en.wikipedia.org/wiki/Fa%C3%A0_di_Bruno%27s_formula):

$$
\frac{\partial^n}{\partial x_1 \dots \partial x_n} =
\sum_{\pi \in \Pi} f^{(|\pi|}(y) \cdot \prod_{B \in \pi} \frac{\partial^{|B|}y}{\prod_{j \in B} \partial x_j}
$$

For $n=2$ the partitions are just $\{\{1\},\{2\}\}$ and $\{\{1,2\}\}$
$$
\frac{\partial^2 f(y)}{\partial x_1 x_2} = 
f'(y) \frac{\partial^2 y}{\partial x_1 x_2} + f''(y) \frac{\partial y}{\partial x_1} \frac{\partial y}{\partial x_2}
$$

The derivative of a matrix cross-product $\mathbf X \mathbf X^T$ should be fairly straightforward (I think?); e.g. for a two-by-two crossproduct works out as $(\mathbf\lambda \mathbf\lambda^T)' = \mathbf\lambda' \mathbf\lambda^T + \mathbf\lambda \mathbf(\lambda^T)'$, where differentiation is element-by-element on the RHS.
